Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex
minor changes in notes from James - keynote section

@@

Arfon Smith of github presented the first keynote. Following points summarize the keynote and the associated discussion:
\begin{enumerate}
    \item Socially driven software development.
    \item Role of software in research.
\end{enumerate}

>>>>

\note{papers and presentations should be references - URLs perhaps can just be footnotes for now, but may want to think about this again.}

@@

\subsection{A recipe for sustainable software, Philip E. Bourne.}

The first keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
was delivered by Philip E. Bourne of UCSD. The slides
are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}

Philip is a basic biomedical scientist but has formed 4 software
companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.

He emphasized that Sustainability for software ``does not just mean more
money from Government.''

He noted that creating effective illustrations is a key function of
software in the biomedical sciences: ``Automating Conceptualization''
(Visualizing molecular structures is a problem that has now been solved,
he is working on 3-d visualizations of the contents of cells.)

He has looked into reproducibility in computational biology, and has
concluded that ``I have proved I cannot reproduce research from my own
lab'':
(Computational
Biology Resources Lack Persistence And Usability, Veretnik, Fink and
Bourne.)\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1000136}} \todo{change this to a cite}

He noted that in places, development in science has improved thanks to
open source and Github, but for the most part remains arcane. He argues
that there's a lot we can learn from the app-store model. He
mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
OpenScienceDataCloud \footnote{\url{https://www.opensciencedatacloud.org/}} as
distribution channels.

He noted a common evolutionary pathway for computational biology
projects, from data archive to analytics platform to educational use. He
suggests that use for outreach might be the final step.

He has problems with staff retention with respect to private organizations which reward 
those combining research and software expertise (the ``Google Bus''), however,
he's a strong fan of private sector sustainability sources and PPP. He
notes that making a successful business from scientific software alone
is very rare: founders overvalue while customers undervalue. He noted
that to last, an open source project needs a minimal funding requirement
even with a vibrant community -- good will only goes so far if you're
being paid to do something else.

He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}} \todo{change at least first to cite}

He had problems with selling research software: his technology transfer
office wanted wanted huge IP reachthrough, whereby they would get a
share of profits from drugs developed by pharma companies who use the
software. He was aware this was unrealistic but the technology transfer
office insisted for a
while. He wants to push a much more one-click approach to purchasing
university-written software.

He then presented the usual arguments on directly valuing software as a
research output alongside papers. Readers will be familiar with this
debate. An interesting reference here is ``What
does peer review mean when applied to computer code?, Rosemary Dickin.''
\footnote{\url{http://blogs.plos.org/biologue/2013/08/08/what-does-peer-review-mean-when-applied-to-computer-code/}} \todo{change to cite}

He then expanded on the idea of the \textbf{digital enterprise}, where
IT underpins the whole of organizational activities. He contends that
universities are way behind the curve on this. In particular, he
highlights the separation of research, teaching and administration into
silos without a common IT framework as a blocker to many useful
organizational innovations: ``University 2.0 is yet to happen.'' He
spoke of a circumstance where someone had used an algorithm developed
for computational biology in marketing.

The role of an institution is important in this space. He argued that
funders can help train institutions not just individuals.

He concluded with a reference to his paper
\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1002001}}``Ten
simple rules for getting ahead as a computational biologist in
academia'' \todo{change to cite} and argued that computational scientists ``have a
responsibility to convince their institutions and communities that
software is scholarship,'' a point with which we strongly agree.

\subsection{Scientific Software and the Open Collaborative Web}

\todo{change references to papers or presentations to cites}

The second keynote\footnote{\url{http://is.gd/wssspe}} was by
 Arfon Smith\footnote{\url{http://arfon.org/}} of Github.

His talk used an example from data reduction in Astronomy, where one
attempts to remove interfering effects from the device. In this field, a
``bad pixel mask'' is recalculated by each researcher, despite being
persistent over time. He estimated 13 wasted person-years of waste.

He asked, of this practice, ``why didn't we know better?'' and argued
this was because we're taught to focus on immediate research outcomes
and not on continuously improving and building on tools for research. He
then asked, later in our working lives, when we do know better, why
don't we act any different, and argued this was due to incentivization:
only the immediate products of research, not the software, are valued.

He referenced Victoria Stodden's talk at OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
which he said argued these points well.

He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov Chain
Monte Carlo code by Dan Forman Mackey\footnote{\url{http://dan.iel.fm/}} as a
good example of a scientific project using GitHub. Amongst other
examples of good scientific projects on GitHub he used an example from Joe Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}.

C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with regard to
reusable software, ``we should just start doing it''. Arfon notes that
documentation should be ``treated as a first class entity.''

He noted that some software fields have excellent cultures of module
reuse, such as rubygems\footnote{\url{http://rubygems.org/}},
PyPI\footnote{\url{https://pypi.python.org/pypi}}, and CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
ultra-low-friction collaboration through these repositories. This has
not happened in highly numerical, compiled language scientific software.

He argues that the GitHub Pull Request code review mechanism facilitates
such collaboration, by allowing one to code first, and seek review and a
merge back into the trunk later.

Fernando Perez says that ``Open Source
is reproducible by necessity'' - since reproducibility is a prerequisite
for remote collaboration.

Arfon therefore defines GitHub as part of the next stage in development
of the web, following on from the ``social web'' of Facebook et.al.:
``The Collaborative Web.''
